---
title: "Variable Importance and Effects"
format:
  html:
    self-contained: true
    embed-resources: true
    df-print: paged
    toc: true
---

# Overview

Diagnostics reflect exactly what scripts/040_downscale.R and scripts/042_downscale_analysis.R produce.

- Variable importance: per-ensemble importance computed during downscaling (040) and summarized by median and IQR across ensembles (042).
- Partial dependence (PDP): top two predictors (by median importance) using the saved RF model and training data from 040.
- ALE (Accumulated Local Effects) and ICE: optional if the iml package is available; computed from the same saved model and training data.

Notes

- Importance units depend on the RF backend: %IncMSE or an equivalent importance metric; higher values indicate stronger influence.
- PDP assumes feature independence; may mislead under strong collinearity. ALE is preferred in that case and is centered at 0.
- The mixed scenario (woody + annual) has no standalone model; therefore VI/PDP/ALE/ICE are not generated for that PFT.
- PDPs are generated from a single saved RF model (first ensemble) for illustration, while importance ranks are aggregated across ensembles.

# Importance + Partial Dependence

```{r, include=FALSE}
options(ccmmf.quiet_banner = TRUE)
source(here::here("000-config.R"))
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}

fig_dir <- here::here("figures")
imp_pngs <- list.files(fig_dir, pattern = "_importance_partial_plots\\.png$", full.names = TRUE)
if (length(imp_pngs) == 0) {
  cat("No importance/partial plots found. Run scripts/042_downscale_analysis.R to generate them.")
} else {
  # Parse filenames into PFT and pool: <pft>_<pool>_importance_partial_plots.png
  meta <- tibble::tibble(file = imp_pngs) |>
    dplyr::mutate(name = basename(file)) |>
    # capture any pool token (last underscore-delimited token before suffix)
    tidyr::extract(name, into = c("pft", "pool"),
                   regex = "^(.*)_([^_]+)_importance_partial_plots\\.png$",
                   remove = FALSE) |>
    dplyr::arrange(pool, pft)

  pools <- unique(meta$pool)
  for (po in pools) {
    cat("\n## ", po, " by PFT\n\n", sep = "")
    mf <- dplyr::filter(meta, pool == po)
    for (i in seq_len(nrow(mf))) {
      cat("### ", gsub("_", " ", mf$pft[i]), "\n\n", sep = "")
      knitr::include_graphics(mf$file[i])
      cat("\n")
      cat("- Left: variable importance (median across ensembles) with interquartile ranges.\n")
      cat("- Middle/Right: PDP for the top two predictors.\n\n")
    }
  }
}
```

# ALE and ICE Effects

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
ale_svgs <- list.files(fig_dir, pattern = "_ALE_predictor[0-9]+\\.svg$", full.names = TRUE)
ice_svgs <- list.files(fig_dir, pattern = "_ICE_predictor[0-9]+\\.svg$", full.names = TRUE)

if (length(ale_svgs) == 0 && length(ice_svgs) == 0) {
  cat("ALE/ICE plots not found (package 'iml' may be missing during generation).\n")
} else {
  parse_fx <- function(files, kind) {
    tibble::tibble(file = files) |>
      dplyr::mutate(name = basename(file)) |>
      # capture any pool token (last underscore-delimited token before kind)
      tidyr::extract(
        name,
        into = c("pft", "pool", "pred_idx"),
        regex = paste0("^(.*)_([^_]+)_", kind, "_predictor([0-9]+)\\.svg$"),
        remove = FALSE
      ) |>
      dplyr::mutate(pred_idx = as.integer(pred_idx))
  }

  ale_meta <- parse_fx(ale_svgs, "ALE")
  ice_meta <- parse_fx(ice_svgs, "ICE")

  # Merge ALE and ICE by pft/pool/pred_idx
  meta <- dplyr::full_join(
    dplyr::rename(ale_meta, ale_file = file),
    dplyr::rename(ice_meta, ice_file = file),
    by = c("pft", "pool", "pred_idx")
  ) |>
    dplyr::arrange(pool, pft, pred_idx)

  pools <- unique(meta$pool)
  for (po in pools) {
    cat("\n## ", po, " – ALE and ICE\n\n", sep = "")
    mf <- dplyr::filter(meta, pool == po)
    pf <- unique(mf$pft)
    for (pf_i in pf) {
      cat("### ", gsub("_", " ", pf_i), "\n\n", sep = "")
      for (k in unique(mf$pred_idx[mf$pft == pf_i])) {
        row <- dplyr::filter(mf, pft == pf_i, pred_idx == k)[1, ]
        cat("#### Top predictor ", k, "\n\n", sep = "")
        if (!is.na(row$ale_file)) knitr::include_graphics(row$ale_file)
        if (!is.na(row$ice_file)) knitr::include_graphics(row$ice_file)
        cat("\n- ALE: centered effect; more robust under collinearity.\n",
            "- ICE: per-observation curves; heterogeneity and interactions appear as diverse or crossing trajectories.\n\n")
      }
    }
  }
}
```
