---
title: "Intro to MAGiC Downscaling"
author:
  - "David LeBauer"
  - "Akash BV"
date: today
quarto-cache: true
format:
  html:
    self-contained: false
    df-print: default
execute:
  cache: true
  echo: true
  warning: false
  message: false
  eval: false
---

# Overview

This document gives a brief overview of the MAGiC downscaling pipeline. 
The downscaling workflow follows the ensemble workflow: it starts from SIPNET NetCDF outputs from a subset of representative design points, uses machine learning to downscale to field-level carbon predictions, and then aggregates to county-level statistics.

```{mermaid}
flowchart LR
  A[PEcAn/SIPNET NetCDF outputs] --> B[Extract to CSV\nscript 030]
  B --> C[Downscale to fields\nscript 040]
  C --> D[Aggregate to county \nscript 041]
```

For this intro we focus on soil carbon (`TotSoilCarb`) for one ensemble member of the baseline scenario in 2024. Other model outputs, including but not limited to aboveground biomass and GHG fluxes (N2O, CH4), can be downscaled and aggregated with the same pipeline.

Pre-requisites for this demo include:
* System setup
    - AWS CLI 
    - A Conda environment
    - The github.com/ccmmf/downscaling repository cloned locally
    - A Linux computing environment
* Input data -- two archives downloaded from S3 storage:
    - Output from SIPNET runs at design points 
    - CADWR Land Use Maps

This tutorial will cover the following components:

1. Workflow Overview: how SIPNET outputs become field-level predictions and county-level aggregates
2. Extract SIPNET output: script 030 converts NetCDF files to `ensemble_output.csv`
3. Downscale to fields: script 040 fits Random Forest models and predicts to fields
4. Inspect outputs: look at tables and quick plots to understand the downscaled predictions
5. Wrap-up: recap and aggregation preview

**Demo Mode** This demo runs a simplified version of the pipeline to focus on the core downscaling steps. The full pipeline includes more variables, scenarios, and ensemble members, which add complexity but follow the same overall structure.

The `--mode demo` flags in the commands below constrain the pipeline to a single slice of the data:
* One variable (TotSoilCarb) is downscaled
* One scenario (baseline), so it does not include scenario-vs-baseline comparisons.
* One ensemble member

Demo mode does not include scenario-vs-baseline comparisons or ensemble based uncertainty analysis.

Outputs:
* `ensemble_output.csv` -- monthly SIPNET output in long format
* `downscaled_preds.csv` -- per-field carbon predictions

# System setup

The pipeline runs inside a conda environment that includes the software required by the downscaling scripts.

The demo data and the `pecan-all` env tarball are downloaded from an S3 bucket hosted at NCSA.

## Check AWS access

First, confirm that the AWS CLI is available:

```bash
aws --version
```

If that command is not found, load the AWS CLI as a module on your HPC system or install it from the [AWS CLI installation instructions](https://docs.aws.amazon.com/cli/latest/userguide/getting-started-install.html).

Next, confirm that you can read the CCMMF S3 bucket:

```bash
aws s3 ls s3://carb/data_raw/
```

If the bucket listing fails because of missing credentials, configure the `ccmmf-garage` profile.

Add this section to `~/.aws/credentials`:

```bash
[ccmmf-garage]
aws_access_key_id = [your key ID]
aws_secret_access_key = [your secret key]
```

Add this section to `~/.aws/config`:

```bash
[profile ccmmf-garage]
region = garage
endpoint_url = https://s3.garage.ccmmf.ncsa.cloud
```

Make sure the credential files are readable only by you:

```bash
chmod 600 ~/.aws/credentials ~/.aws/config
```

Activate the profile for the current shell and check access again:

```bash
export AWS_PROFILE=ccmmf-garage
aws s3 ls s3://carb/data_raw/
```

## Setup Conda environment

We will use a conda environment that provides R, PEcAn, and all of the R packages the pipeline scripts depend on. The environment is available as a tarball on S3. 

In this example, ~/.conda/envs/pecan-all is the target location, but you may want to put this somewhere other than your home directory.

Note that this may take 30 minutes or more to download and install.

```bash
aws s3 cp s3://carb/deploy/setup-pecan-env.sh ./
bash setup-pecan-env.sh 1.10 ~/.conda/envs/pecan-all
```

To activate the pecan-all environment in a new shell:

```bash
conda activate ~/.conda/envs/pecan-all
```

## Clone the downscaling repository

The pipeline scripts live in the downscaling repo, which is located at https://github.com/ccmmf/downscaling. Clone the repository to your working directory -- the rest of the steps below assume you're inside the cloned directory.

```bash
git clone -b develop https://github.com/ccmmf/downscaling.git
cd downscaling
```

## Input data

We install two data archives from S3. The first is the demo subset of SIPNET outputs (baseline scenario, ensemble 1, year 2024). The second is a vector file of California crop fields and county geometry layers derived from the CADWR Land Use Dataset [Cite].

```bash
mkdir -p ~/ccmmf_demo/modelout ~/ccmmf_demo/data

aws s3 cp s3://carb/data_raw/ccmmf_phase_3_scenarios_v2_n2o_ch4_demo.tgz - | \
  tar xzf - -C ~/ccmmf_demo/modelout/

aws s3 cp s3://carb/data_raw/ccmmf_downscaling_data_layers.tgz - | \
  tar xzf - -C ~/ccmmf_demo/data/
```

Note: to run a full production run, you can download `ccmmf_phase_3_scenarios_v2_n2o_ch4.tgz`. This is about 95 GB and may take 30 minutes or longer to download.

Verify the layout:

```bash
ls ~/ccmmf_demo/modelout/ccmmf_phase_3_scenarios_v2_n2o_ch4/
ls ~/ccmmf_demo/data/
```

In `modelout/...` you should see `output_baseline/` and `site_info.csv`. In `data/` you should see `ca_fields.gpkg`, `ca_field_attributes.csv`, `site_covariates.csv`, and `ca_counties.gpkg`.

These are the PEcAn model output files that script 030 will extract:

```bash
find ~/ccmmf_demo/modelout/ccmmf_phase_3_scenarios_v2_n2o_ch4/output_baseline/out \
  -name "2024.nc" | head
```

What does one of these files contain?

```bash
ncdump -h ~/ccmmf_demo/modelout/ccmmf_phase_3_scenarios_v2_n2o_ch4/output_baseline/out/ENS-00001-2eab840584d9a6ea/2024.nc
```

Show the covariate table that script 040 will use:

```bash
head -n 5 ~/ccmmf_demo/data/site_covariates.csv | column -s, -t
```

Set the `CCMMF_DIR` environment variable so that the pipeline knows where to find the data:

```bash
export CCMMF_DIR=~/ccmmf_demo
```

This tells the pipeline where to find `$CCMMF_DIR/modelout/...` and `$CCMMF_DIR/data/...`, as well as where to write pipeline outputs.

# Extract SIPNET output

The `030_extract_sipnet_output.R` script reads the PEcAn standard netCDF model output, aggregates to monthly values, and writes these values to a long CSV, `ensemble_output.csv`.

Let's run it in demo mode:

```bash
Rscript scripts/030_extract_sipnet_output.R --mode demo
```

Confirm and inspect the output:

```bash
ls -lh ~/ccmmf_demo/modelout/ccmmf_phase_3_scenarios_v2_n2o_ch4/ensemble_output.csv
```

```{r inspect-ensemble-output}
model_outdir <- "~/ccmmf_demo/modelout/ccmmf_phase_3_scenarios_v2_n2o_ch4"
ensemble_output <- readr::read_csv(file.path(model_outdir, "ensemble_output.csv"), 
  show_col_types = FALSE)
dplyr::glimpse(ensemble_output)
ensemble_output |>
  dplyr::summarise(
    rows = dplyr::n(),
    sites = dplyr::n_distinct(site_id),
    ensembles = dplyr::n_distinct(parameter),
    variables = paste(unique(variable), collapse = ", "),
    start = min(datetime),
    end = max(datetime)
  )
```

# Downscaling from Model Outputs to fields

Now that we have extracted the model outputs to a tabular format, we can use these outputs to train machine learning models that predict field-level carbon stocks and fluxes.

This happens in `scripts/040_downscale.R`. 

For each run, `scripts/040_downscale.R`:

1. Loads `ensemble_output.csv`, site covariates, and field attributes.
2. Iterates over each `(scenario, PFT, carbon pool)` combination.
3. Fits one Random Forest model per ensemble member for that combination.
4. Predicts to fields for that PFT.
5. Writes outputs including `downscaled_preds.csv` (and `.parquet` for large outputs).

Fitting one RF model per ensemble member preserves and propagates uncertainty structure throughout the full modeling workflow.

```bash
Rscript scripts/040_downscale.R --mode demo
```

In demo mode there is one (scenario, PFT, pool) combo and one ensemble member and it predicts to roughly 10,000 target fields within a couple of minutes.

Outputs are written to `$CCMMF_DIR/modelout/.../downscaled_preds.csv` (and `.parquet`).

# Inspect downscaled outputs

```bash
ls -lh ~/ccmmf_demo/modelout/ccmmf_phase_3_scenarios_v2_n2o_ch4/downscaled_preds.csv
```

```r
model_outdir <- "~/ccmmf_demo/modelout/ccmmf_phase_3_scenarios_v2_n2o_ch4"
preds <- readr::read_csv(file.path(model_outdir, "downscaled_preds.csv"), show_col_types = FALSE)
dplyr::glimpse(preds)
preds |>
  dplyr::summarise(
    rows = dplyr::n(),
    fields = dplyr::n_distinct(site_id),
    counties = dplyr::n_distinct(county),
    min_density = min(density_per_ha, na.rm = TRUE),
    mean_density = mean(density_per_ha, na.rm = TRUE),
    max_density = max(density_per_ha, na.rm = TRUE)
  )
```

Focus on `density_per_ha`, `scenario`, `pft`, `ensemble`, `county`, and `model_output`. The `area_ha` column is useful later for converting density to total field carbon, but it is not the main explanatory variable for this first demo.

# Make quick plots

Use `density_per_ha` for the first plot because it is the direct field-level downscaling result.

```{r plot-downscaled-density}
library(dplyr)
library(ggplot2)
library(readr)

model_outdir <- "~/ccmmf_demo/modelout/ccmmf_phase_3_scenarios_v2_n2o_ch4"
preds <- read_csv(file.path(model_outdir, "downscaled_preds.csv"), show_col_types = FALSE)

ggplot(preds, aes(x = density_per_ha)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  labs(
    title = "Downscaled Total Soil Carbon",
    x = "Density (Mg C / ha)",
    y = "Number of fields"
  ) +
  theme_minimal()
```

Join predictions to covariates to show how predicted soil carbon varies with a predictor such as organic carbon density (`ocd`) or temperature (`temp`).

```{r plot-density-by-covariates}
covariates <- read_csv("~/ccmmf_demo/data/site_covariates.csv", show_col_types = FALSE)

plot_data <- preds |>
  left_join(
    covariates |> select(site_id, ocd, temp, vapr, clay, precip, srad, twi),
    by = "site_id"
  ) |>
  filter(!is.na(ocd), !is.na(density_per_ha))

plot_sample <- plot_data |>
  slice_sample(n = min(nrow(plot_data), 5000))

ggplot(plot_sample, aes(x = ocd, y = density_per_ha)) +
  geom_point(alpha = 0.25, size = 0.7) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE, color = "black") +
  labs(
    title = "Predicted Soil Carbon vs. Soil Organic Carbon Density",
    x = "SoilGrids organic carbon density covariate",
    y = "Predicted density (Mg C / ha)"
  ) +
  theme_minimal()
```

# Aggregate to county

This section is still under construction, will be presented in a follow-up demo.

The `041_aggregate_to_county.R` script computes per-county totals from field-level predictions. This is a downstream aggregation step and is intentionally skipped in the first live demo.

```bash
# Rscript scripts/041_aggregate_to_county.R --mode demo
```

# Per-ensemble metrics and variable importance (script 042, follow-up)

```bash
# Rscript scripts/042_downscale_analysis.R --mode demo
```

Computes per-ensemble downscaling metrics (R^2, RMSE), plus accumulated local effect (ALE) and partial dependence (PDP) plots for the top covariates.

Outputs in `figures/`:

* `variable_importance_summary.png`
* per-spec ALE / ICE SVG files

# County maps and scenario deltas (script 043, follow-up)

```bash
# Rscript scripts/043_county_level_plots.R --mode demo
```

State-level maps plus per-county delta layers. The delta layers show how soil carbon changed between the start and end of the simulation period.

# Where to go next

* **Production mode**: drop `--mode demo` (production is the default) and run on the full archive. Plan for several hours.
* **Other variables**: AGB, N2O flux, CH4 flux. Same pipeline, different reporting units -- Mg/ha for stocks, kg/ha/yr for fluxes.
* **Atlas reports**: `reports/atlas_*.qmd` render long-form HTML reports off these outputs.
* **New scenario**: extend `management_scenarios` in `000-config.R` and re-run.
